\input{stuff/config.tex}
\newcommand{\detrz}{\rc{əte}\slash{}\obj{e-}}
\usepackage{phonrule}
\usepackage{amsthm,etoolbox,xcolor}

\title{Irregular first person markers in Cariban\\}

\author{Florian Matter \\ University of Bern}
%\date{}

\begin{document}
\pagestyle{empty}
\maketitle	

\begin{abstract}
\PC is reconstructed as having an verbal inflectional paradigm called \dbqu{\setone} \parencite{gildea1998}.
Most modern languages have reflexes of this system, which shows hierarchically conditioned access to index slots in transitive verbs, and two lexically conditioned inflection classes, \gl{s_a_} and \gl{s_p_}, in intransitive verbs.
This split-\gl{s} division is not based on any known semantic or pragmatic criteria, but rather, the overwhelming majority of \gl{s_a_} verbs are synchronically or diachronically derived from transitive verbs by means of a detransitivizing prefix \parencite{meira2000split}, which had the forms \detrz{} \parencite{meira2010origin}.
In many languages, person markers in this system have been subject to innovations, often by the extension of person markers from somewhere in paradigm to a new location.
In intransitive verbs, this often lead to the erosion or even loss of the split-\gl{s} system.
These person marker extensions happened by way of lexical diffusion, spreading within the verbal lexicon.

Six of these extensions did not fully run their course, but left a small group of verbs with old markers (i.e., irregularly inflected verbs).
This is illustrated in \cref{tab:hixka} for \hixka (Parukotoan), where the regular first person prefix is \obj{k(ɨ)-}, but the verb \obj{to} \qu{to go} shows an irregular first person marker \obj{ɨ-}, since it was not affected by the extension of innovative \obj{k(ɨ)-}.
%
\begin{table}[b]
\caption{\hixka \gl{s_a_} verbs \parencites{hixkaryanaderby1985}{howard2001wrought} }
\label{tab:hixka}
\centering
\begin{tabular}{@{}lllll@{}}
\mytoprule
{} &  \qu{to be standing} &     \qu{to fall} &  \qu{to be afraid} &    \qu{to go} \\
\mymidrule
\gl{1}   &  \obj{k-osowaʃehto-} &  \obj{k-ehurka-} &  \obj{k-oserʲehɨ-} &   \obj{ɨ-to-} \\
\gl{2}   &  \obj{m-osowaʃehto-} &  \obj{m-ehurka-} &  \obj{m-oserʲehɨ-} &  \obj{mɨ-to-} \\
\gl{1+2} &  \obj{t-osowaʃehto-} &  \obj{t-ehurka-} &  \obj{t-oserʲehɨ-} &  \obj{tɨ-to-} \\
\gl{3}   &  \obj{n-osowaʃehto-} &  \obj{ɲ-ehurka-} &  \obj{n-oserʲehɨ-} &   \obj{n-to-} \\
\mybottomrule
\end{tabular}
\end{table}
%
Interestingly, these incomplete extensions, three of which can be reconstructed to proto-languages, all introduced new first person \gl{s_a_} markers.
In addition, the \gl{s_a_} verbs which resisted the extensions show a considerable etymological overlap across the different cases of extension.

I reconstruct these verbs, which often show other irregularities, and investigate potential reasons for why these particular verbs were not affected by different extensions.
Using \posscite{bybee1985morphology} network model of morphology, I show that some cases of incomplete extension can be argued to be due to lexical connections based on morphology (presence vs absence of the detransitivizing prefix), while others appear to be due to phonological connections (\obj{e}- and/or \obj{ə}-initial vs others).
While semantic connections appear to play no role, many of the resistant verbs are high frequency verbs, a factor predicting conservativism in Bybee’s model.
In fact, most cases of resistant verbs are predicted by all three factors: morphological connections, phonological connections, and frequency.
Thus, while the network model offers attractive explanations for the (non-)spread of innovative \gl{1}\gl{s_a_} markers, these predicting factors strongly overlap in many of the investigated cases of extension.
This in turn is largely due to the fact that the most high-frequency \gl{s_a_} verbs were exceptional already at the level of \PC, since they did not contain the detransitivizing prefix \detrz.

\printbibliography

\end{abstract}

\end{document}